Longitudinal Principal Manifold Estimation

### 2023-12-14 paper/main.tex

The framework for principal manifolds originated with the concept of principal curves (PS, \cite{hastiePrincipalCurves1989}), which are essentially curves that pass through the middle of a data cloud. However, as pointed out by \cite{duchamp1996extremal}, the PS framework in \cite{hastiePrincipalCurves1989} has limitations such as the model complexity selection. In addition, applying the methods in high-dimensional settings requires is problematic. Motivated by the penalization approaches presented by in \cite{kegl2000learning} and \cite{smolaRegularizedPrincipalManifolds2001}, \cite{mengPrincipalManifoldEstimation2021} proposed the principal manifold estimation (PME) framework. In order to present the longitudinal PME approach, we start by introducing notations used throughout this paper. We use $d$ and $D$ to denote the dimensions of the low-dimensional manifold and high-dimensional space, respectively. We assume that a data cloud is observed in the $D$ dimensional space. In the context of segmentation, we consider a collection of vertices on the surface of the brain region of interest as the $D-$dimensional data cloud, while the $d-$dimensional smooth surface of the region is the underlying manifold of interest. For any positive integer $q$ and point $\xi=(\xi_1,\ldots,\xi_q)\in\mathbb{R}^q$, we use the following norm $\Vert \xi\Vert_{\mathbb{R}^q}=\sqrt{\sum_{l=1}^q \xi_l^2}$. Since our interest in this work centers at continuous mappings, we denote the collection of continuous vector-valued functions from $\mathbb{R}^d$ to $\mathbb{R}^D$ as $\mathcal{C}(\mathbb{R}^{d} \to \mathbb{R}^{D})$. The $d$-dimensional manifold corresponding to a function $f \in C(\mathbb{R}^{d} \to \mathbb{R}^{D})$ is formally defined as $M_f^d=\{f(r):\,r\in\mathbb{R}^d\}$. Most manifold estimation methods are based on defining a projection index, i.e. $\pi_f(x)$, defined as the point in the $d$-dimensional space, where $f(\pi_f(x))$ is the projection of $x$ on the manifold $M_f^d$. The definitions of $\pi_f(x)$ given by \cite{hastiePrincipalCurves1989} and \cite{mengPrincipalManifoldEstimation2021} for $2$- and higher-dimensional spaces, respectively, guarantee the uniqueness of the projection. The principal manifolds are then defined as follows. 

@@

The framework described above is an excellent approach for fitting a 2-dimensional principal surface that passes through the ``middle'' of a 3-dimensional cloud of vertices. We first heuristically explain the goal of our proposed model and its importance in segmentation of brain regions for longitudinal imaging studies. In other words, we consider the problem of reducing the dimension of the coordinates $(x_1, x_2, x_3)$, i.e. $D=3$ to $(y_1,y_2)$ with $d=2$, which are the projections of the original coordinates on the surface of the 3-dimensional shape of interest. Now consider that the data are observed longitudinally, during several visits by a study participant. Hence, the data of interest consists of the 4-dimensional cloud $(x_1, x_2, x_3,t_j)$, where $t_j$ denotes the time of the $j$th study visit. Figure \ref{fig:sample_data} displays a simplified example of this setting using simulated data where $D = 2$.

\begin{figure}[h]
  \centering
  \includegraphics[width=\textwidth]{sim_case1_plot2}
  \caption{Longitudinal Simulated Data with $D = 2$}
  \label{fig:sample_data}
\end{figure}

@@

\begin{figure}%
  \centering
  \subfloat[\centering Data]{{\includegraphics[width=9cm]{sim_case1_plot2}}}%
  \hfill
  \subfloat[\centering Step 1: Sample Size Reduction]{{\includegraphics[width=9cm]{lpme_data_reduction}}}
  \vfill
  \subfloat[\centering Step 2: Initialization]{{\includegraphics[width=9cm]{lpme_initialization}}}
  \hfill
  \subfloat[\centering Steps 3 and 4: Fitting and Tuning]{{\includegraphics[width=9cm]{lpme_fitting}}}
  \caption{LPME Algorithm Steps}
  \label{fig:lpme_steps}
\end{figure}

@@

Given that in most manifold learning tasks in practice data are often quite large, we begin by implementing an approach for selecting a small number number of points in $D$-dimensional space that represent the structure of the data cloud. This approach is aimed at reducing the sample size of the original data, while maintaining the available information on the curvature of the underlying manifold of interest. Specifically, we consider $D$-dimensional observations $\left\{x_{it}\right\}_{i=1, t=1}^{I_t, T}$ where the sample size $I_t$ is assumed to be large. To reduce the sample size $I_t$, we use the HDMDE method (see \cite{mengPrincipalManifoldEstimation2021}, Section 5.3) to estimate a reduced set of points in the $D$-dimensional space that represent the shape of the data cloud $\left\{x_{it}\right\}_{i=1, t=1}^{I_t, T}$. Briefly, this approach is based on clustering of the original time points to obtain cluster centers, $\left\{\mu_{j, t}\right\}_{j=1, t=1}^{N_t, T}$, along with their corresponding weights $\{\hat{\theta}_{j, t}\}_{j=1, t=1}^{N_t, T}$ that provide necessary information to characterize the shape of the original data. The number of clusters is estimated for each time point and is allowed to vary among time points to accommodate potential changes in the shape of the underlying structure. By implementing this approach, we reduce the size $\sum_{t=1}^T I_t$ of data $\left\{x_{it}\right\}_{i=1, t=1}^{I_t, T}$ from $I_t$ to $\sum_{t=1}^T N_t$, where each $N_t$ tends to be much smaller than the corresponding $I_t$ (see the simulation study presented in Figure 3(c) of \cite{mengPrincipalManifoldEstimation2021}). The centers estimated in this step are then used to reach an initial low-dimensional parameterization.

@@

The next step in developing the algorithm for estimation of the longitudinal principal manifold is initialization. Different algorithms use different approaches for this step. For example, the PME algorithm uses Isomap (\cite{tenenbaumGlobalGeometricFramework2000}) to develop an initial parameterization, which the algorithm proceeds to iteratively improve. However, implementation of commonly used approaches for initialization in our setting can be problematic. For instance, using ISOMAP to find parameterizations of similarly shaped structures (i.e. the brain region of interest at each time point) is likely to result in inconsistent parameterization directions (i.e. low values in the parameter space may be associated with low values in the embedding space for one run, while low values in the parameter space may be associated with high values in the embedding space in another). This is demonstrated in Figure \ref{fig:inconsistent_parameterization}. Reaching a successful fit in the LPME algorithm requires consistent parameterizations across time points, so a different approach to generating an initial parameterization is needed.

\begin{figure}[h]
  \centering
  \includegraphics[width=\textwidth]{inconsistent_parameterization}
  \caption{Simulated Data with Inconsistent Parameterization}
  \label{fig:inconsistent_parameterization}
\end{figure}

To obtain the same parameterization of the manifold over time, instead of using Isomap on the data for each time point separately, we propose collapsing the time dimension in the cluster centers of all time steps, then using one Isomap run to reach a parameterization for the centers across all time points simultaneously. This approach is based on the assumption that the shape of the structure, i.e. the brain region of interest, remains relatively stable over time. In Section \ref{s:simulations}, we consider some deviations from this assumption and their effect on the results and we show that this process ensures that the initial parameterization remains consistent over time under the assumption of relative stability of the shape of the structure.

@@

Once the initial parameterization for each cluster center is obtained, the model fitting process begins by using this parameterization as the initialization for runs of the PME algorithm at each time point. Specifically, for each time point, the cluster centers associated with the given time value, their weights, and the corresponding parameterizations for these points are used to initialize a PME run. This results in an estimated principal manifold $\widehat{f_t}$ for each time point $t$ and an optimal tuning parameter $\lambda_t^*$ corresponding to the first two terms of the $\mathcal{K}_{\lambda, \gamma, \mathbb{P}}(F)$ defined in ~\eqref{eq:newKappa}. At this stage, the time-specific manifolds require further smoothing considering the curvature in the time dimension corresponding to the third term of $\mathcal{K}_{\lambda, \gamma, \mathbb{P}}(F)$. The principal manifold $\widehat{f_t}$ at time $t$ is characterized by spline coefficients. Hence, we estimate a spline function mapping from the time dimension to the coefficient space that smooths over the coefficients for these principal manifolds, given a tuning parameter $\gamma$.

@@

Once longitudinal estimates $F_{\gamma}(t, \boldsymbol{r})$ have been estimated for all provided tuning values, the optimal tuning parameter $\gamma^*$ is identified using leave-one-out cross validation. In this process, the coefficient smoothing function is computed while excluding all data and coefficients with time $t$ from consideration, for $t = 1, \dots, T$. We denote this function by $g_{\gamma}^{(t)}$. We can then assess the performance of the LPME model using tuning value $\gamma$ by averaging over estimates of MSD calculated using the original observations associated with the left-out time $t$ as follows.
\begin{equation}
  MSD(\gamma) = \frac{1}{T} \sum_{t=1}^{T}\frac{1}{I_t}\sum_{i=1}^{I_t}\|x_{i, t} - f_{\gamma}^{(t)}(t, \pi_{f_{\gamma}^{(t)}}(x_{i, t}))\|^2. \label{eq:18}
\end{equation}
The optimal tuning value $\gamma^*$ is then found to be the value of $\gamma$ that minimizes $MSD(\gamma)$. Note that alternative cross validation methods, such as $k$-fold cross validation, could easily be used in place of leave-one-out cross validation to better meet computational demands.

@@

@@

\begin{figure}%
  \centering
  \subfloat[\centering Unaugmented Unit Circle]{{\includegraphics[width=9cm]{unit_circle_D2}}}%
  \hfill
  \subfloat[\centering Augmented Unit Circle]{{\includegraphics[width=9cm]{unit_circle_D3}}}
  \caption{Illustration of our proposed data augmentation approach using the Unit Circle. The left panel (a) shows the depiction of the circle with $d=1$ and $D=2$, while the right panel (b) shows how our proposed use of considering this same manifold in $D=3$ by using the spherical coordinates results in a manifold that is not self-intersecting.}
  \label{fig:unit_circle_augmentation}
\end{figure}

@@

\begin{table}
\small
  \centering
  \begin{tabular}{|c c c c c|}
    \hline
    Case & $d$ & $D$ & $f_t(\mathbf{r})$ & Domain \\
    \hline
    1 & 1 & 2 & $\left(r, \ \alpha \text{sin} \ (\beta r + \frac{\pi}{2})\right) + \zeta\mathbf{g}(t) + \iota$ & $-3 \leq r \leq 3$ \\
    2 & 1 & 2 & $\left(r, \ \alpha\text{sin} \ (\beta r)\right) + \zeta\mathbf{g}(t) + \iota$ & $-3\pi \leq r \leq 3\pi$ \\
    3 & 1 & 2 & $\left(\alpha_1 \text{cos} \ (\beta_1 r), \ \alpha_2\text{cos} \ (\beta_2 r)\right) + \zeta\mathbf{g}(t) + \iota$ & $-\frac{4\pi}{5} \leq r \leq \frac{\pi}{2}$ \\
    4 & 1 & 3 & $\left(r, (\alpha_1r + \beta_1)^2, \ (\alpha_2r + \beta_2)^3\right) + \zeta\mathbf{g}(t) + \iota$ & $-1 \leq r \leq 1$ \\
    5 & 1 & 3 & $\left(r, \ \alpha_1\text{cos} \ (\beta_1 r), \ \alpha_2\text{sin} \ (\beta_2 r) \right) + \zeta\mathbf{g}(t) + \iota$ & $0 \leq r \leq 3\pi$ \\
    6 & 2 & 3 & $\left(\beta_1r_1, \ \beta_2r_2, \ \alpha_1(\alpha_2\|\beta\mathbf{r}\|^2)\right) + \zeta\mathbf{g}(t) + \iota$ & $-1 \leq r_1, r_2 \leq 1$\\
    7 & 2 & 3 & $\left(\alpha_1\beta_1r_1\text{cos} \ (\alpha_1r_1), \ \alpha_2\beta_2r_1\text{sin} \ (\alpha_2r_1), \ r_2\right) + \zeta\mathbf{g}(t) + \iota$ & $0 \leq r_1 \leq 3\pi$; \ $-1 \leq r_2 \leq 1$\\
    8 & 2 & 3 & $\left(\alpha_1\text{sin}(\beta_1r_1)\text{cos}(\beta_2r_2), \ \alpha_1\text{sin}(\beta_1r_1)\text{sin}(\beta_2r_2), \ \alpha_1\text{cos}(\beta_1r_1)\right) + \zeta\mathbf{g}(t) + \iota$ & $0 \leq r_1 \leq \pi$; \ $0 \leq r_2 \leq 2\pi$\\
    \hline
  \end{tabular}
  \caption{Embedding functions used for simulation trials.}
  \label{table:simulation_embeddings}
\end{table}

@@

A factorial design is used to run the simulation studies, with the factor levels set as follows:
\begin{itemize}
  \item $\alpha, \beta, \zeta \in \left\{0, 0.05, 0.1, 0.25, 0.5, 1.0\right\}$
  \item Study duration: $\left\{1, 2, 5\right\}$
  \item Interval between images: $\left\{0.1, 0.25, 0.5\right\}$
  \item Longitudinal change model: Constant, Linear, Quadratic, Sinusoidal
\end{itemize}
The sample size for each time point is set to 1,000 observations, so each embedding map is simulated a total of 7,776 times. All code required to reproduce the simulation results and the associated figures is provided at \href{https://github.com/rjzielinski/lpme-project}{\texttt{https://github.com/rjzielinski/lpme-project}}.

@@

\begin{table}[h]
  \centering
  \begin{tabular}{|c c c c c|}
    \hline
    Case & Data & LPME & PME & PC/PS \\
    \hline
    1 & 0.223 (0.256) & 0.125 (0.161) & 0.268 (0.850) & 0.189 (0.245) \\
    2 & 0.514 (0.408) & 0.384 (0.648) & 0.843 (1.93) & 0.600 (0.296) \\
    3 & 0.446 (0.445) & 0.401 (0.446) & 0.507 (0.594) & 0.412 (0.423) \\
    4 & 30.7 (88.1) & 27.7 (260) & 30.7 (88.2) & 30.6 (88.1) \\
    5 & 0.980 (0.771) & 0.791 (0.845) & 1.04 (1.07) & 0.934 (0.713) \\
    6 & 1.43 (6.04) & 1.21 (5.66) & 1.47 (6.06) & 1.01 (2.11) \\
    7 & 0.580 (0.839) & 4.07 (3.30) & 7.37 (1.14) & 1.95 (0.800) \\
    8 & 0.226 (0.275) & 0.136 (0.169) & 0.242 (0.311) & 0.274 (0.243) \\
    \hline
  \end{tabular}
  \caption{MSD Comparison to True Values, Mean (SD).}
  \label{table:simulation_results_mean}
\end{table}

@@

\begin{figure}[h]
  \centering
  \includegraphics[width=\textwidth]{sim_case1}
  \caption{Simulation Case 1}
  \label{fig:sim_case1}
\end{figure}

\begin{figure}[h]
  \centering
  \includegraphics[width=\textwidth]{sim_case5}
  \caption{Simulation Case 5}
  \label{fig:sim_case4}
\end{figure}

@@

\zielinski{It is key to note here that the MSD values shown in these tables are comparing the various models' projections onto the estimated manifold to the true values underlying the data the models are fit on. This means that a model that performs well using this metric may not perform well when comparing the data to the model's projections onto the manifold. In fact, if there is a high level of inter-observation noise, then we should expect an estimated manifold that closely follows the true underlying manifold to show a relatively poor fit to the data at any given time point.

This largely explains why PME does not usually show better performance than the PC/PS methods using this metric - both approaches are trying to estimate a curve or surface using the data available at a given time point, which may have noise introduced that increases the distance to the true manifold. When comparing the fit of the PME and PC methods to the simulated data, PME tends to show improvement in the median MSD, with elevated mean MSD levels compared to principal curves.}

@@

Images were processed using FSL and ANTs via the \texttt{fslr} and \texttt{ANTsR} packages in \texttt{R}. Specifically, ANTs was used for bias correcting the images, and the FIRST method in FSL was used for image segmentation. Following image segmentation, the surfaces of each hippocampus were identified by finding the extreme voxels in each dimension with nonzero intensity readings. The estimated hippocampus surface positions were then standardized to a maximum distance of one from the origin in Euclidean space and centered around the origin. Finally, the data were augmented with spherical coordinates in a manner similar to that described in the previous section to avoid the need to fit to a self-intersecting manifold.

\zielinski{Reaching volume estimates from the LPME output is proving to be difficult.}

@@

\zielinski{Figure \ref{fig:adni_result} will continue to be improved, with the current version mainly serving as a placeholder.}

@@

@@

\meng{FYI, we may also apply Gaussian process regression (GPR) to smooth the time direction. The smoothness can be represented by the covariance kernel we use in the GPR. For example, we may apply the Gaussian or \href{https://en.wikipedia.org/wiki/Matern_covariance_function}{Matérn} kernels. The taste of the GPR is similar to \cite{dunsonInferringManifoldsNoisy2022}, and we may cite \cite{dunsonInferringManifoldsNoisy2022}.}
\newpage

>>>>

The framework for principal manifolds originated with the concept of principal curves (PS, \cite{hastiePrincipalCurves1989}), which are essentially curves that pass through the middle of a data cloud. However, as pointed out by \cite{duchamp1996extremal}, the PS framework in \cite{hastiePrincipalCurves1989} has limitations such as the model complexity selection. In addition, applying the methods in high-dimensional settings requires is problematic. Motivated by the penalization approaches presented by \cite{kegl2000learning} and \cite{smolaRegularizedPrincipalManifolds2001}, \cite{mengPrincipalManifoldEstimation2021} proposed the principal manifold estimation (PME) framework. In order to present the longitudinal PME approach, we start by introducing notations used throughout this paper. We use $d$ and $D$ to denote the dimensions of the low-dimensional manifold and high-dimensional space, respectively. We assume that a data cloud is observed in the $D$ dimensional space. In the context of segmentation, we consider a collection of vertices on the surface of the brain region of interest as the $D-$dimensional data cloud, while the $d-$dimensional smooth surface of the region is the underlying manifold of interest. For any positive integer $q$ and point $\xi=(\xi_1,\ldots,\xi_q)\in\mathbb{R}^q$, we use the following norm $\Vert \xi\Vert_{\mathbb{R}^q}=\sqrt{\sum_{l=1}^q \xi_l^2}$. Since our interest in this work centers at continuous mappings, we denote the collection of continuous vector-valued functions from $\mathbb{R}^d$ to $\mathbb{R}^D$ as $\mathcal{C}(\mathbb{R}^{d} \to \mathbb{R}^{D})$. The $d$-dimensional manifold corresponding to a function $f \in C(\mathbb{R}^{d} \to \mathbb{R}^{D})$ is formally defined as $M_f^d=\{f(r):\,r\in\mathbb{R}^d\}$. Most manifold estimation methods are based on defining a projection index, i.e. $\pi_f(x)$, defined as the point in the $d$-dimensional space, where $f(\pi_f(x))$ is the projection of $x$ on the manifold $M_f^d$. The definitions of $\pi_f(x)$ given by \cite{hastiePrincipalCurves1989} and \cite{mengPrincipalManifoldEstimation2021} for $2$- and higher-dimensional spaces, respectively, guarantee the uniqueness of the projection. The principal manifolds are then defined as follows. 

@@

The framework described above is an excellent approach for fitting a 2-dimensional principal surface that passes through the ``middle'' of a 3-dimensional cloud of vertices. We first heuristically explain the goal of our proposed model and its importance in segmentation of brain regions for longitudinal imaging studies. In other words, we consider the problem of reducing the dimension of the coordinates $(x_1, x_2, x_3)$, i.e. $D=3$ to $(y_1,y_2)$ with $d=2$, which are the projections of the original coordinates on the surface of the 3-dimensional shape of interest. Now consider that the data are observed longitudinally, during several visits by a study participant. Hence, the data of interest consists of the 4-dimensional cloud $(x_1, x_2, x_3,t_j)$, where $t_j$ denotes the time of the $j$th study visit. Figure \ref{fig:sample_data} displays a simplified example of this setting using simulated data where $D = 2$. The goal is to fit a surface through the data that depends on time and is not affected by the random noise added at each time point and in space.

\begin{figure}[h]
  \centering
  \includegraphics[scale=0.4]{sim_case1_plot2}
  \caption{Longitudinal Simulated Data with $D = 2$. Data from the function ADD FUNCTION are simulated at 5 time points. Random noise is added both in time and space as ADD FUNCTION.}
  \label{fig:sample_data}
\end{figure}

@@

\begin{figure}[ht]
  \centering
  \subfloat[\centering Data]{{\includegraphics[width=6.5cm]{sim_case1_plot2}}}%
  \hfill
  \subfloat[\centering Step 1: Sample Size Reduction]{{\includegraphics[width=6.5cm]{lpme_data_reduction}}}
  \vfill
  \subfloat[\centering Step 2: Initialization]{{\includegraphics[width=6.5cm]{lpme_initialization}}}
  \hfill
  \subfloat[\centering Steps 3 and 4: Fitting and Tuning]{{\includegraphics[width=6.5cm]{lpme_fitting}}}
  \caption{LPME Algorithm Steps. In (a) we show the original data cloud, in (b) we show the sample size reduction step, where the red points indicate the representative data to be used for the next steps of the algorithm, (c) shows the fitted time specific curves, the blue surface in (d) shows the estimated smooth surface. }
  \label{fig:lpme_steps}
\end{figure}

@@

Given that in most manifold learning tasks in practice data are often quite large, we begin by implementing an approach for selecting a small number number of points in $D$-dimensional space that represent the structure of the data cloud. This approach is aimed at reducing the sample size of the original data, while maintaining the available information on the curvature of the underlying manifold of interest. Specifically, we consider $D$-dimensional observations $\left\{x_{it}\right\}_{i=1, t=1}^{I_t, T}$ where the sample size $I_t$ is assumed to be large. To reduce the sample size $I_t$, we use the HDMDE method (see \cite{mengPrincipalManifoldEstimation2021}, Section 5.3) to estimate a reduced set of points in the $D$-dimensional space that represent the shape of the data cloud $\left\{x_{it}\right\}_{i=1, t=1}^{I_t, T}$. Briefly, this approach is based on clustering of the original time points to obtain cluster centers, $\left\{\mu_{j, t}\right\}_{j=1, t=1}^{N_t, T}$, represented using red dots in Figure \ref{fig:lpme_steps} (b), along with their corresponding weights $\{\hat{\theta}_{j, t}\}_{j=1, t=1}^{N_t, T}$ that provide necessary information to characterize the shape of the original data. The number of clusters is estimated for each time point and is allowed to vary among time points to accommodate potential changes in the shape of the underlying structure. By implementing this approach, we reduce the size $\sum_{t=1}^T I_t$ of data $\left\{x_{it}\right\}_{i=1, t=1}^{I_t, T}$ from $I_t$ to $\sum_{t=1}^T N_t$, where each $N_t$ tends to be much smaller than the corresponding $I_t$ (see the simulation study presented in Figure 3(c) of \cite{mengPrincipalManifoldEstimation2021}). The centers estimated in this step are then used to reach an initial low-dimensional parameterization.

@@

The next step in developing the algorithm for estimation of the longitudinal principal manifold is initialization. Different algorithms use different approaches for this step. For example, the PME algorithm uses Isomap (\cite{tenenbaumGlobalGeometricFramework2000}) to develop an initial parameterization, which the algorithm proceeds to iteratively improve. However, implementation of commonly used approaches for initialization in our setting can be problematic. For instance, using ISOMAP to find parameterizations of similarly shaped structures (i.e. the brain region of interest at each time point) is likely to result in inconsistent parameterization directions (i.e. low values in the parameter space may be associated with low values in the embedding space for one run, while low values in the parameter space may be associated with high values in the embedding space in another). This is demonstrated in Figure \ref{fig:inconsistent_parameterization} which shows differences in parameterizations depending on the time points of the data cloud presented in Figure \ref{fig:sample_data}. Reaching a successful fit in the LPME algorithm requires consistent parameterizations across time points, so a different approach to generating an initial parameterization is needed.

\begin{figure}[h]
  \centering
  \includegraphics[scale=0.6]{inconsistent_parameterization}
  \caption{Simulated Data with Inconsistent Parameterization}
  \label{fig:inconsistent_parameterization}
\end{figure}

To obtain the same parameterization of the manifold over time, instead of using Isomap at each time point separately, we propose collapsing the time dimension in the cluster centers of all time steps, then using one Isomap run to reach a parameterization for the centers across all time points simultaneously. This approach is based on the assumption that the shape of the structure, i.e. the brain region of interest, remains relatively stable over time. In Section \ref{s:simulations}, we consider some deviations from this assumption and their effect on the results and we show that this process ensures that the initial parameterization remains consistent over time under the assumption of relative stability of the shape of the structure.

@@

Once the initial parameterization for each cluster center is obtained, the model fitting process begins by using this parameterization as the initialization for runs of the PME algorithm at each time point. Specifically, for each time point, the cluster centers associated with the given time value, their weights, and the corresponding parameterizations for these points are used to initialize a PME run. This results in an estimated principal manifold $\widehat{f_t}$ for each time point $t$ and an optimal tuning parameter $\lambda_t^*$ corresponding to the first two terms of the $\mathcal{K}_{\lambda, \gamma, \mathbb{P}}(F)$ defined in ~\eqref{eq:newKappa} and presented using multicolored lines in Figure \ref{fig:lpme_steps} (c). At this stage, the time-specific manifolds require further smoothing considering the curvature in the time dimension corresponding to the third term of $\mathcal{K}_{\lambda, \gamma, \mathbb{P}}(F)$. The principal manifold $\widehat{f_t}$ at time $t$ is characterized by spline coefficients. Hence, we estimate a spline function mapping from the time dimension to the coefficient space that smooths over the coefficients for these principal manifolds, given a tuning parameter $\gamma$.

@@

Once $F_{\gamma}(t, \boldsymbol{r})$ have been estimated for all provided tuning values, the optimal tuning parameter $\gamma^*$ is identified using leave-one-out cross validation. In this process, the coefficient smoothing function is computed while excluding all data and coefficients with time $t$ from consideration, for $t = 1, \dots, T$. We denote this function by $g_{\gamma}^{(t)}$. We can then assess the performance of the LPME model using tuning value $\gamma$ by averaging over estimates of MSD calculated using the original observations associated with the left-out time $t$ as follows.
\begin{equation}
  MSD(\gamma) = \frac{1}{T} \sum_{t=1}^{T}\frac{1}{I_t}\sum_{i=1}^{I_t}\|x_{i, t} - f_{\gamma}^{(t)}(t, \pi_{f_{\gamma}^{(t)}}(x_{i, t}))\|^2. \label{eq:18}
\end{equation}
The optimal tuning value $\gamma^*$ is then found to be the value of $\gamma$ that minimizes $MSD(\gamma)$. Note that alternative cross validation methods, such as $k$-fold cross validation, could easily be used in place of leave-one-out cross validation to better meet computational demands.

@@

\begin{figure}[ht]
  \centering
  \subfloat[\centering Unaugmented Unit Circle]{{\includegraphics[width=9cm]{unit_circle_D2}}}%
  \hfill
  \subfloat[\centering Augmented Unit Circle]{{\includegraphics[width=9cm]{unit_circle_D3}}}
  \caption{Illustration of our proposed data augmentation approach using the Unit Circle. The left panel (a) shows the depiction of the circle with $d=1$ and $D=2$, while the right panel (b) shows how our proposed use of considering this same manifold in $D=3$ by using the spherical coordinates results in a manifold that is not self-intersecting.}
  \label{fig:unit_circle_augmentation}
\end{figure}

@@

@@

\begin{table}[ht]
\small
  \centering
  \begin{tabular}{|c c c c c|}
    \hline
    Case & $d$ & $D$ & $f_t(\mathbf{r})$ & Domain \\
    \hline
    1 & 1 & 2 & $\left(r, \ \alpha \text{sin} \ (\beta r + \frac{\pi}{2})\right) + \zeta\mathbf{g}(t) + \iota$ & $-3 \leq r \leq 3$ \\
    2 & 1 & 2 & $\left(r, \ \alpha\text{sin} \ (\beta r)\right) + \zeta\mathbf{g}(t) + \iota$ & $-3\pi \leq r \leq 3\pi$ \\
    3 & 1 & 2 & $\left(\alpha_1 \text{cos} \ (\beta_1 r), \ \alpha_2\text{cos} \ (\beta_2 r)\right) + \zeta\mathbf{g}(t) + \iota$ & $-\frac{4\pi}{5} \leq r \leq \frac{\pi}{2}$ \\
    4 & 1 & 3 & $\left(r, (\alpha_1r + \beta_1)^2, \ (\alpha_2r + \beta_2)^3\right) + \zeta\mathbf{g}(t) + \iota$ & $-1 \leq r \leq 1$ \\
    5 & 1 & 3 & $\left(r, \ \alpha_1\text{cos} \ (\beta_1 r), \ \alpha_2\text{sin} \ (\beta_2 r) \right) + \zeta\mathbf{g}(t) + \iota$ & $0 \leq r \leq 3\pi$ \\
    6 & 2 & 3 & $\left(\beta_1r_1, \ \beta_2r_2, \ \alpha_1(\alpha_2\|\beta\mathbf{r}\|^2)\right) + \zeta\mathbf{g}(t) + \iota$ & $-1 \leq r_1, r_2 \leq 1$\\
    7 & 2 & 3 & $\left(\alpha_1\beta_1r_1\text{cos} \ (\alpha_1r_1), \ \alpha_2\beta_2r_1\text{sin} \ (\alpha_2r_1), \ r_2\right) + \zeta\mathbf{g}(t) + \iota$ & $0 \leq r_1 \leq 3\pi$; \ $-1 \leq r_2 \leq 1$\\
    8 & 2 & 3 & $\left(\alpha_1\text{sin}(\beta_1r_1)\text{cos}(\beta_2r_2), \ \alpha_1\text{sin}(\beta_1r_1)\text{sin}(\beta_2r_2), \ \alpha_1\text{cos}(\beta_1r_1)\right) + \zeta\mathbf{g}(t) + \iota$ & $0 \leq r_1 \leq \pi$; \ $0 \leq r_2 \leq 2\pi$\\
    \hline
  \end{tabular}
  \caption{Embedding functions used for simulation trials.}
  \label{table:simulation_embeddings}
\end{table}

@@

A factorial design is used to run the simulation studies, with the factor levels set as follows: 1) $\alpha, \beta, \zeta \in \left\{0, 0.05, 0.1, 0.25, 0.5, 1.0\right\}$, 2) study duration: $\left\{1, 2, 5\right\}$, 3) interval between images: $\left\{0.1, 0.25, 0.5\right\}$, 4) longitudinal change model: Constant, Linear, Quadratic, Sinusoidal. The sample size for each time point is set to 1,000 observations, so each embedding map is simulated a total of 7,776 times. All code required to reproduce the simulation results and the associated figures is provided at \href{https://github.com/rjzielinski/lpme-project}{\texttt{https://github.com/rjzielinski/lpme-project}}.

@@

\begin{table}[h]
  \centering
  \begin{tabular}{|c c c c c|}
    \hline
    Case & Data & LPME & PME & PC/PS \\
    \hline
    1 & 0.223 (0.256) & {\bf 0.125 (0.161)} & 0.268 (0.850) & 0.189 (0.245) \\
    2 & 0.514 (0.408) & {\bf 0.384 (0.648)} & 0.843 (1.93) & 0.600 (0.296) \\
    3 & 0.446 (0.445) & {\bf 0.401 (0.446)} & 0.507 (0.594) & 0.412 (0.423) \\
    4 & 30.7 (88.1) & {\bf 27.7 (260)} & 30.7 (88.2) & 30.6 (88.1) \\
    5 & 0.980 (0.771) & {\bf 0.791 (0.845)} & 1.04 (1.07) & 0.934 (0.713) \\
    6 & 1.43 (6.04) & 1.21 (5.66) & 1.47 (6.06) & {\bf 1.01 (2.11)} \\
    7 & 0.580 (0.839) & 4.07 (3.30) & 7.37 (1.14) & {\bf 1.95 (0.800)} \\
    8 & 0.226 (0.275) & {\bf 0.136 (0.169)} & 0.242 (0.311) & 0.274 (0.243) \\
    \hline
  \end{tabular}
  \caption{MSD comparison to true values, Mean (SD). For each case, the lowest algorithm-specific mean (SD) are highlighted in bold. }
  \label{table:simulation_results_mean}
\end{table}

@@

\begin{figure}[h]
  \centering
  \includegraphics[width=\textwidth]{sim_case1}
  \includegraphics[width=\textwidth]{sim_case5}
  \caption{Simulation Cases 1 and 5.}
  \label{fig:sim_case1}
\end{figure}

@@

@@

Images were processed using FSL and ANTs via the \texttt{fslr} and \texttt{ANTsR} packages in \texttt{R}. Specifically, ANTs was used for bias correcting the images, and the FIRST method in FSL was used for image segmentation. Following image segmentation, the surfaces of each hippocampus were identified by finding the extreme voxels in each dimension with nonzero intensity readings. The estimated hippocampus surface positions were then standardized to a maximum distance of one from the origin in Euclidean space and centered around the origin. Finally, the data were augmented with spherical coordinates in a manner similar to that described in Section \ref{s:LPME} to avoid the need to fit to a self-intersecting manifold.

%\zielinski{Reaching volume estimates from the LPME output is proving to be difficult.}

@@

%\zielinski{Figure \ref{fig:adni_result} will continue to be improved, with the current version mainly serving as a placeholder.}

@@

It is key to note here that the MSD values shown in these tables are comparing the various models' projections onto the estimated manifold to the true values underlying the data the models are fit on. This means that a model that performs well using this metric may not perform well when comparing the data to the model's projections onto the manifold. In fact, if there is a high level of inter-observation noise, then we should expect an estimated manifold that closely follows the true underlying manifold to show a relatively poor fit to the data at any given time point. This largely explains why PME does not usually show better performance than the PC/PS methods using this metric - both approaches are trying to estimate a curve or surface using the data available at a given time point, which may have noise introduced that increases the distance to the true manifold. When comparing the fit of the PME and PC methods to the simulated data, PME tends to show improvement in the median MSD, with elevated mean MSD levels compared to principal curves.

@@

%\meng{FYI, we may also apply Gaussian process regression (GPR) to smooth the time direction. The smoothness can be represented by the covariance kernel we use in the GPR. For example, we may apply the Gaussian or \href{https://en.wikipedia.org/wiki/Matern_covariance_function}{Matérn} kernels. The taste of the GPR is similar to \cite{dunsonInferringManifoldsNoisy2022}, and we may cite \cite{dunsonInferringManifoldsNoisy2022}.}
\newpage
